% ====================================================================
% === SUPERSEDED (reader-facing) by Docs/notes/07_calibration_objective.tex
%     (Vol-Fitter Technical Notes series). For the maintained,
%     comprehensive treatment with implementation + performance
%     appendices and code-generated figures, read that note.
% === RETAINED here as the equation-label citation source for code
%     docstrings, which reference this file by path and by its eq.
%     labels (e.g. "eq. q_logit", "note section 5.2").
% ====================================================================
\documentclass[9pt,a4paper]{extarticle}

\usepackage[margin=0.55in]{geometry}
\usepackage{amsmath,amssymb,booktabs}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{listings}
\usepackage{titlesec}

\setlength{\parindent}{0pt}
\setlength{\parskip}{2.5pt}
\setlength{\abovedisplayskip}{4pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{3pt}
\setlength{\belowdisplayshortskip}{3pt}
\titlespacing*{\section}{0pt}{0.65em}{0.25em}
\lstset{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  frame=single,
  breaklines=true,
  tabsize=4,
  xleftmargin=0pt,
  xrightmargin=0pt
}

\title{Fast Time-Value Density Weights for Implied-Volatility Fitting}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-2.6em}

\section*{Objective}
Let
\[
  x_i=\log(K_i/F)
\]
be normalized strike. Each quote is treated as at-the-money or out-of-the-money, so its observed option-price level is its time value, denoted $\mathrm{TV}_i>0$.

The desired benchmark rule is: if quotes lie on the uniform grid
\[
  x_k=-2+\frac{4k}{n},\qquad k=0,\ldots,n,
\]
then each quote receives weight
\[
  w_k=\mathrm{TV}_k.
\]
For non-uniform market strikes, the objective is different from merely assigning $w_i=\mathrm{TV}_i$. If quotes are dense in one region and sparse elsewhere, raw time-value weights overweight the dense region simply because it has more observations. The desired aggregate distribution of weights over normalized-strike space should still have the shape of time value.

Equivalently, the target continuous weighting measure is
\[
  d\mu(x)\propto \mathrm{TV}(x)\,dx.
\]
For a strike interval $A$, total fitting importance should therefore scale like
\[
  \mu(A)\propto \int_A \mathrm{TV}(x)\,dx.
\]
If the local quote density is $\rho(x)$, the naive weights $w_i=\mathrm{TV}_i$ instead approximate
\[
  \mathrm{TV}(x)\rho(x)\,dx,
\]
which mixes the desired economic weight with the accidental sampling density. Hence one should divide by local quote density:
\[
  w_i\propto \frac{\mathrm{TV}_i}{\rho(x_i)}.
\]
In one dimension, $1/\rho(x_i)$ is estimated by a local strike spacing.

\section*{Weight construction}
Sort the quotes by normalized strike:
\[
  x_1<x_2<\cdots<x_m.
\]
Define local cell widths
\[
  s_1=x_2-x_1,\qquad
  s_i=\frac{x_{i+1}-x_{i-1}}{2}\quad(2\le i\le m-1),\qquad
  s_m=x_m-x_{m-1}.
\]
This is a one-dimensional Voronoi-cell approximation. Let
\[
  \bar{s}=\frac{1}{m}\sum_{i=1}^m s_i.
\]
The proposed weight is
\[
  \boxed{w_i=\mathrm{TV}_i\frac{s_i}{\bar{s}}.}
\]
Dense regions have small $s_i$ and are downweighted; sparse regions have large $s_i$ and are upweighted. The normalization by $\bar{s}$ keeps the scale comparable to raw time-value weights and gives the exact benchmark property on uniform grids.

In a least-squares fit with model parameters $\theta$, minimize
\[
  \min_\theta \sum_{i=1}^m w_i r_i(\theta)^2.
\]
For implied-volatility residuals,
\[
  r_i(\theta)=\sigma_{\mathrm{model}}(x_i;\theta)-\sigma_{\mathrm{obs},i}.
\]
If an optimizer expects ordinary residuals, pass $\sqrt{w_i}r_i(\theta)$.

\section*{Uniform-grid check}
On the grid $x_k=-2+4k/n$, every spacing is $h=4/n$. Therefore $s_k=h$ for all $k$, and $\bar{s}=h$. The rule reduces to
\[
  w_k=\mathrm{TV}_k\frac{h}{h}=\mathrm{TV}_k.
\]
Thus the non-uniform correction is inactive when quotes are already uniformly distributed in normalized-strike space.

\section*{Example}
Consider five quotes:
\[
  x=(-0.50,-0.10,0.00,0.05,0.80),\qquad
  \mathrm{TV}=(0.10,0.40,0.50,0.45,0.08).
\]
The middle quotes are dense, while the right wing is sparse. The local spacings are
\[
  s=(0.40,0.25,0.075,0.40,0.75),\qquad \bar{s}=0.375.
\]
Hence the spacing multipliers are
\[
  \frac{s}{\bar{s}}\approx(1.067,0.667,0.200,1.067,2.000),
\]
and the final weights are
\[
  w\approx(0.107,0.267,0.100,0.480,0.160).
\]
The quote at $x=0$ has large time value but lies in a crowded region, so its individual weight is reduced. The isolated right-wing quote has low time value but represents a wider strike interval. Aggregated over strike bins, the weights follow a time-value-shaped density rather than the raw quote-count histogram.

\section*{Python implementation}
\begin{lstlisting}
import numpy as np

def tv_density_weights(x, tv, eps=1e-12, max_mult=None):
    """Density-corrected time-value weights for IV fitting."""
    x = np.asarray(x, dtype=float)
    tv = np.asarray(tv, dtype=float)
    if x.shape != tv.shape:
        raise ValueError("x and tv must have the same shape")

    order = np.argsort(x)
    xs = x[order]
    tvs = np.maximum(tv[order], eps)
    m = len(xs)

    if m == 0:
        return np.array([], dtype=float)
    if m == 1:
        weights_sorted = np.array([tvs[0]], dtype=float)
    else:
        s = np.empty(m)
        s[0] = xs[1] - xs[0]
        s[-1] = xs[-1] - xs[-2]
        if m > 2:
            s[1:-1] = 0.5 * (xs[2:] - xs[:-2])
        s = np.maximum(s, eps)

        mult = s / np.mean(s)
        if max_mult is not None:
            mult = np.minimum(mult, max_mult)
        weights_sorted = tvs * mult

    weights = np.empty_like(weights_sorted)
    weights[order] = weights_sorted
    return weights

# scipy.optimize.least_squares usage:
# weights = tv_density_weights(x, tv, max_mult=10.0)
# def residuals(theta):
#     return np.sqrt(weights) * (model_iv(theta, x) - sigma_obs)
\end{lstlisting}

\section*{Practical notes}
If several quotes have the same normalized strike, aggregate them first or split the strike-level weight among duplicates. Otherwise duplicate quotes mechanically overweight that strike. In production, use
\[
  w_i=\max(\mathrm{TV}_i,\varepsilon)\frac{s_i}{\bar{s}},
\]
and optionally cap $s_i/\bar{s}$, for example at $5$ or $10$, to prevent a single isolated far-wing quote from dominating the fit.

\end{document}
