\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{volfit_preamble}
\graphicspath{{figures/}}

\title{\textbf{Observation Kalman Filtering}\\[2pt]
\large Denoising observed smiles without confusing noise with gaps\\[2pt]
\normalsize Vol-Fitter Technical Notes, No.~15}
\author{Vol-Fitter}
\date{\today}

% Note-local notation.
\newcommand{\obs}{\mathrm{obs}}
\newcommand{\pred}{\mathrm{pred}}
\newcommand{\post}{\mathrm{post}}
\newcommand{\gap}{\mathrm{gap}}
\newcommand{\calL}{\mathcal L}
\newcommand{\calI}{\mathcal I}
\newcommand{\calO}{\mathcal O}
\newcommand{\calQ}{\mathcal Q}
\newcommand{\calR}{\mathcal R}
\newcommand{\calH}{\mathcal H}

\begin{document}
\maketitle

\begin{abstract}
Prior persistence and Kalman filtering both pull a smile away from a raw,
current-market fit, so they are easy to confuse in implementation. They should
not be confused mathematically. Prior persistence is a \emph{gap regularizer}: it
keeps unobserved or weakly identified smile functionals near a transported prior
and turns itself off where today's quotes already identify the functional. A
Kalman observation filter is a \emph{temporal state estimator}: it assumes the
current smile is observed through noisy, sometimes contradictory quotes, predicts
the smile from the previous filtered state, and updates according to the
relative precision of the prediction and the observation. This note proposes a
clean Vol-Fitter architecture for adding such a filter. The state is not raw
option mids, but compact arbitrage-safe smile handles; the transition is the
existing SSR transport; the measurement is a data-only fit with a covariance
derived from bid--ask width, quote density, fit residuals and local
inconsistency; and active mode is a one-stage MAP calibration so the same quotes
are not counted twice. A small case file shows the key distinction: dense but
contradictory close strikes should inflate observation noise and be denoised by
the Kalman layer, while genuinely unquoted wings should be governed by prior
persistence.
\end{abstract}

\tableofcontents
\newpage

%======================================================================
\section{The production symptom}
\label{sec:symptom}

A sparse wing and a noisy cluster are different failures.

In a sparse wing, the desk has almost no market information. A data-only fit can
extend the ATM skew into the tail and create a new signal where no option
actually traded. Note~13 solved that problem with an activation gate: the prior
is strong in the gap and exactly off where the data is sufficient.

In a noisy cluster, the desk has information, but it is internally inconsistent.
Two adjacent strikes may imply a local kink because one market is stale, a wide
bid--ask band may make the mid meaningless, or an American wing repair may have
preserved spread while lowering the reliable rank of the wing. The right answer
is not "fill with yesterday"; the right answer is "infer the latent current
smile under observation noise."

\begin{invariant}
\begin{enumerate}[leftmargin=1.6em]
\item Prior persistence acts only on functionals not identified by the current
market at the required precision.
\item Observation filtering acts only through an explicit prediction covariance
and an explicit measurement covariance.
\item The same quote cannot be used once to create a filtered posterior and then
again as an independent calibration datum.
\item The filter state is an arbitrage-safe coordinate, such as ATM handles or
operator values, never a free raw-IV curve.
\item Every filtered output reports prediction, observation, innovation, gain
and posterior uncertainty.
\end{enumerate}
\end{invariant}

The two features therefore answer two different questions:
\[
\begin{array}{ll}
\text{Prior persistence:} & \text{Where do we lack enough observations to
identify the smile?}\\[2pt]
\text{Kalman filtering:} & \text{Given observations here, how noisy are they?}
\end{array}
\]
The implementation should preserve that distinction all the way to the UI.

%======================================================================
\section{Two regularizers, two measures}
\label{sec:split}

Let $x_t\in\R^d$ denote the latent smile state of one node at snapshot time
$t$: for the first implementation this should be a compact vector of handles,
for example
\[
x_t=(\sigma_{\atm,t},s_{\atm,t},\kappa_{\atm,t})\T,
\]
possibly extended later by signed operators such as RR25, BF25 and var-swap.
Let $q_t$ be the prepared quote set after forwards, de-Americanization,
event-clock conversion, bid--ask band construction and quote edits.

\subsection{Prior persistence}

Prior persistence is a penalty in a calibration objective. With a prior state
$x_t^0$ transported to the current forward, and functionals
$O_j:\R^d\to\R$, Note~13 has the universal form
\begin{equation}
\calL_{\mathrm{persist}}(x)
=\half\sum_j
\lambda_j^{\gap}
\left(\frac{O_j(x)-O_j(x_t^0)}{a_j}\right)^2,
\qquad
\lambda_j^{\gap}
=\lambda_j^0
\max\!\left(1-\frac{\pi^{\obs}_j}{\pi^{\mathrm{req}}_j},0\right)^\gamma .
\label{eq:persistence}
\end{equation}
The gate is driven by \emph{identifiability}: when current quotes identify
functional $O_j$ at or above the required precision, $\lambda_j^{\gap}=0$.
Thus prior persistence is not primarily a denoiser. It is a rule for refusing
to invent signal outside the support of the observation.

\subsection{Observation filtering}

A Kalman filter is instead a sequential estimator. It starts from the previous
posterior law for $x_{t-1}$, transports it forward, then combines the transported
prediction with today's noisy observation. In the linear-Gaussian handle model,
\begin{align}
x_t &= \Phi_t x_{t-1}+u_t+\eta_t,
&\eta_t&\sim\mathcal N(0,Q_t), \label{eq:state}\\
z_t &= H_t x_t+\epsilon_t,
&\epsilon_t&\sim\mathcal N(0,R_t), \label{eq:measurement}
\end{align}
where $z_t$ is not a raw quote vector; it is a noisy handle estimate extracted
from today's prepared quotes. $Q_t$ is process covariance: time elapsed, spot
move, event risk and transport distance. $R_t$ is measurement covariance:
bid--ask width, quote density, stale fields, de-Americanization repairs, band
slack and contradictory close strikes.

\begin{heuristic}
Persistence says "do not put signal where the market did not speak." Filtering
says "when the market speaks with a noisy voice, infer what it probably meant."
Both can pull toward a previous surface, but the sufficient statistic is
different: persistence uses a \emph{gap}; filtering uses a \emph{covariance}.
\end{heuristic}

%======================================================================
\section{The Kalman object}
\label{sec:kalman}

The Vol-Fitter should implement the filter in covariance form, matching the
graph posterior style of Note~14. For a direct handle observation $H_t=I$, the
form is especially transparent, but the equations below allow a subset or a
linear operator measurement.

\begin{definition}[Predicted and observed handle laws]
\label{def:laws}
At snapshot $t$, the prediction law and the measurement law are
\[
x_t\given\mathcal F_{t-1}\sim\mathcal N(m_t^-,P_t^-),
\qquad
z_t\given x_t\sim\mathcal N(H_t x_t,R_t).
\]
$m_t^-$ is the transported previous filtered state; $P_t^-$ is its transported
uncertainty plus process noise. $z_t$ and $R_t$ are extracted from today's
prepared quotes without using prior persistence.
\end{definition}

The central update is
\begin{equation}
\boxed{\;
\begin{aligned}
S_t&=H_tP_t^-H_t\T+R_t,\\
K_t&=P_t^-H_t\T S_t^{-1},\\
m_t^+&=m_t^-+K_t\big(z_t-H_tm_t^-\big),\\
P_t^+&=(I-K_tH_t)P_t^-(I-K_tH_t)\T+K_tR_tK_t\T .
\end{aligned}
\;}
\label{eq:kalman}
\end{equation}
The last line is the Joseph form; it is a little longer than
$P_t^+=(I-K_tH_t)P_t^-$, but it preserves symmetry and positive
semidefiniteness better in floating point.

\begin{proposition}[Precision-weighted shrinkage]
\label{prop:scalar}
In one dimension with $H_t=1$, prediction variance $p$ and measurement variance
$r$, the posterior mean is
\[
m^+=\frac{r}{p+r}m^-+\frac{p}{p+r}z,
\]
and the Kalman gain is $K=p/(p+r)$.
\end{proposition}

\begin{proof}
The innovation variance is $S=p+r$, so $K=p/S$. Substituting in
$m^+=m^-+K(z-m^-)$ gives
$m^+=(1-K)m^-+Kz=r(p+r)^{-1}m^-+p(p+r)^{-1}z$.
\end{proof}

\Cref{prop:scalar} is the whole desk intuition. If today's observation is tight
($r\ll p$), the filtered smile moves to the market. If today's observation is
noisy ($r\gg p$), it stays near the transported prediction. The movement is not
controlled by whether the region is observed; it is controlled by the relative
uncertainties of prediction and observation.

%======================================================================
\section{Where the observation covariance comes from}
\label{sec:measurement}

The hard part is not the Kalman algebra. The hard part is building a
measurement covariance $R_t$ that reflects the quote set honestly.

\subsection{Do not filter raw mids}

Raw contracts are a poor state space. The OTM side changes when the forward
moves; contracts appear and disappear; American contracts need early-exercise
removal; a strike can be inside a wide spread but far from its midpoint; and a
raw implied-vol vector is not guaranteed to remain butterfly-free after
smoothing. The filter should begin after the existing quote-prep pipeline in
\nolinkurl{backend/volfit/api/quotes.py}.

\subsection{Extract a noisy handle observation}

The recommended v1 is a two-step observation extractor:
\[
z_t=\argmin_x\ \calL_{\mathrm{quote}}(x;q_t)+\calL_{\mathrm{intrinsic}}(x),
\label{eq:extract}
\]
where $\calL_{\mathrm{quote}}$ is the same vega-normalized/bid--ask-aware loss
as Note~07, and $\calL_{\mathrm{intrinsic}}$ contains only intrinsic model
regularization needed to obtain a valid no-arbitrage smile. It must not include
temporal prior persistence. The extractor then maps the fitted smile to handles
$z_t=(\hat\sigma_{\atm},\hat s_{\atm},\hat\kappa_{\atm})$.

Near the optimum, linearize the quote residuals as
\[
r(\theta)\approx r(\hat\theta)+J(\theta-\hat\theta).
\]
If the handle map is $x=g(\theta)$ with Jacobian $G=\partial g/\partial\theta$,
then an observed-information approximation is
\begin{equation}
\calI_\theta=J\T WJ+\Lambda_{\mathrm{intrinsic}},
\qquad
R_x\approx G\,\calI_\theta\pinv G\T .
\label{eq:cov-delta}
\end{equation}
Equivalently, if the model is already parameterized by handles, use the handle
Hessian directly. In either case the covariance is inflated by realized
inconsistency:
\begin{equation}
R_t=\rho_t R_x,\qquad
\rho_t=\max\left(1,\frac{\chi_t^2}{\max(m-d,1)}\right),
\qquad
\chi_t^2=r(\hat\theta)\T W r(\hat\theta).
\label{eq:resid-inflation}
\end{equation}
This is the mechanism that makes close-strike contradictions visible to the
filter. A contradictory cluster may have many quotes, but if it cannot be fitted
within its stated noise, $\rho_t$ grows and the Kalman gain falls in the
directions the cluster does not reliably identify.

\begin{remark}[Bands and covariance]
For bid--ask mode, quotes inside the band should not pretend to identify the
mid. In a practical implementation, inactive hinge rows contribute little or
nothing to $J\T WJ$, while the weak mid anchor contributes only at its small
anchor weight. This matches Note~07's semantics: inside the spread the market is
a set, not a point.
\end{remark}

\subsection{A cheaper v1 precision builder}

The full covariance \eqref{eq:cov-delta} is the right target, but a first
implementation can reuse the existing precision vocabulary:
\[
R_t^{-1}
=\diag(c_h)\,
\pi_{\mathrm{fit}}\,
f_{\mathrm{density}}\,
f_{\mathrm{spread}}\,
f_{\mathrm{fresh}}\,
f_{\mathrm{repair}}\,
f_{\mathrm{resid}},
\label{eq:cheapR}
\]
where $c_h$ is per-handle confidence, $\pi_{\mathrm{fit}}$ is inverse squared
RMS with a floor and cap, and the scalar factors are exactly the kind already
used by \nolinkurl{backend/volfit/calib/precision.py} and
\nolinkurl{backend/volfit/graph/precision.py}. This is less elegant than a full
Jacobian covariance but enough for an overlay pilot.

%======================================================================
\section{The prediction law}
\label{sec:prediction}

The prediction must transport the previous filtered smile before comparing it
with today's quotes. For a handle state,
\[
m_t^-=\mathcal T_t(m_{t-1}^+),
\qquad
P_t^-=A_tP_{t-1}^+A_t\T+Q_t,
\label{eq:transport}
\]
where $\mathcal T_t$ is the SSR/spot-vol transport of Note~12 and
$A_t=\partial\mathcal T_t/\partial x$ is its local Jacobian. For a small v1,
$A_t=I$ is acceptable for ATM/skew/curvature if the state is recomputed after
transport; the important point is that process covariance $Q_t$ grows when the
prediction is less trustworthy.

Useful process-noise components are
\[
Q_t=Q_{\mathrm{clock}}(\Delta t)+Q_{\mathrm{spot}}(\abs h)
+Q_{\mathrm{event}}+Q_{\mathrm{source}}+Q_{\mathrm{model}},
\label{eq:Q}
\]
with the following meanings:
\begin{description}[leftmargin=1.7em,style=nextline]
\item[Clock noise.]
Volatility surfaces diffuse through time. A square-root-time scale,
$q_h^2\Delta t$ per handle, is the natural baseline.
\item[Spot transport noise.]
The larger the log-forward move $h=\log(F_t/F_{t-1})$, the less certain the
transported state. This is the same intuition as the transport-distance factor
in prior persistence.
\item[Event noise.]
Known event windows should widen the prediction before the event is resolved.
The event clock changes the time variable; this term admits that the shape
response itself is uncertain.
\item[Source noise.]
Switching provider, as-of mode, or live/previous-close source should widen or
reset the filter. A live quote stream and a prior-close snapshot are not the
same stochastic clock.
\item[Model noise.]
If the display model changes from LQD to SVI or Multi-Core SIV, the filtered
handle state may persist, but model-specific parameters should not.
\end{description}

\begin{caution}
Do not filter native local-vol parameters. The local-vol grid is a projection
machinery with its own PDE stability and no-arbitrage constraints. Filter the
target smile handles, then project or refit the local-vol surface to that target.
\end{caution}

%======================================================================
\section{The clean architecture}
\label{sec:architecture}

There are two mathematically safe ways to make the filter active.

\subsection{Overlay mode: compute, display, do not affect calibration}

Overlay mode is the recommended first delivery. The workflow is:
\begin{enumerate}[leftmargin=1.6em]
\item Fetch or stream an option snapshot.
\item Prepare quotes through the existing pipeline.
\item Run the current data-only calibration, with prior persistence off for the
measurement pass.
\item Extract $z_t$ and $R_t$.
\item Predict $(m_t^-,P_t^-)$ from the previous filtered state.
\item Apply \eqref{eq:kalman}; store $(m_t^+,P_t^+)$ and diagnostics.
\item Draw raw calibration, transported prediction and filtered handles/curve.
\end{enumerate}
This mode cannot double-count quotes because it does not feed the posterior back
into the fit. It is therefore ideal for tuning $Q_t$, $R_t$ and reset rules.

\subsection{Active mode: one-stage MAP, not posterior plus quotes}

Active mode should not first build $m_t^+$ from today's quotes and then add
$m_t^+$ as a prior while fitting the same quotes again. That counts $q_t$ twice.
The correct active objective is the one-step MAP problem
\begin{equation}
\calL_{\mathrm{active}}(\theta)
=\calL_{\mathrm{quote}}(\theta;q_t)
+\half\norm[\Big]{\calH(\theta)-m_t^-}_{(P_t^-)^{-1}}^2
+\calL_{\mathrm{intrinsic}}(\theta)
+\calL_{\mathrm{persist},\perp}(\theta).
\label{eq:active-map}
\end{equation}
Here $\calH(\theta)$ extracts the filtered handle coordinates from the model.
The second term is the Kalman prediction prior. The quote likelihood is today's
market. In the linear-Gaussian case, minimizing \eqref{eq:active-map} gives the
same $m_t^+$ as \eqref{eq:kalman}. In the nonlinear smile case it is the extended
Kalman/MAP version.

$\calL_{\mathrm{persist},\perp}$ is the remaining prior-persistence term, but
only on coordinates not already covered by the filter's prediction prior or on
regions outside the filtered handle state. This is the clean split:
\[
\begin{aligned}
\text{Kalman term}
&=\text{temporal prior on observed latent state},\\
\text{persistence term}
&=\text{gap prior on unobserved functionals}.
\end{aligned}
\]

\begin{proposition}[No double counting in the MAP form]
\label{prop:nodouble}
If the model is linear, the quote extractor is $z=H x+\epsilon$ with
$\epsilon\sim\mathcal N(0,R)$, and the prediction is
$x\sim\mathcal N(m^-,P^-)$, then the minimizer of
\[
\half\norm{z-Hx}_{R^{-1}}^2+\half\norm{x-m^-}_{(P^-)^{-1}}^2
\]
is the Kalman posterior mean in \eqref{eq:kalman}.
\end{proposition}

\begin{proof}
The first-order condition is
\[
\big(H\T R^{-1}H+(P^-)^{-1}\big)x
=H\T R^{-1}z+(P^-)^{-1}m^- .
\]
The Woodbury identity rewrites the inverse of the left-hand side as
$P^- - P^-H\T(HP^-H\T+R)^{-1}HP^-$. Substitution gives
$x=m^-+P^-H\T(HP^-H\T+R)^{-1}(z-Hm^-)$, which is \eqref{eq:kalman}.
\end{proof}

\subsection{Where prior persistence remains}

Prior persistence should remain available in active mode, but with explicit
semantics:
\begin{itemize}[leftmargin=1.4em]
\item It may act on deep tails or operator baskets not represented in the
Kalman state.
\item It may provide the initial saved prior from which the first filtered state
is seeded.
\item It may act on dark graph nodes that have no current quote observation.
\item It should not also anchor ATM/skew/curvature to the same previous state
that already appears as $m_t^-$ in the Kalman term.
\end{itemize}

%======================================================================
\section{A telling case file}
\label{sec:case}

\begin{example}[title={Case file: close-strike contradiction versus true gap}]
\textbf{Setup.} A one-month equity smile has nine near-ATM quotes and no
reliable quotes beyond $25\Delta$. The previous filtered state transported to
today is
\[
m^-=(20.0\%,\ -0.35,\ 0.10),\qquad
\sqrt{\diag P^-}=(0.30\%,\ 0.08,\ 0.05).
\]
The near-ATM quotes imply a data-only handle observation
\[
z=(20.4\%,\ -0.37,\ 0.55).
\]
The ATM level and skew are consistent, but one stale close strike forces a very
large curvature if the fit chases mids literally.

\textbf{Diagnosis.} Quote density alone would say the region is well observed.
Therefore prior persistence should switch off for ATM/skew/curvature; using it
to pull curvature back would violate the Note~13 rule. The contradiction should
instead appear as measurement noise. Suppose the observation standard deviations
after residual inflation are
\[
\sqrt{\diag R}=(0.15\%,\ 0.05,\ 0.30).
\]
The scalar gains are approximately
\[
K_{\sigma}=\frac{0.30^2}{0.30^2+0.15^2}=0.80,\quad
K_s=\frac{0.08^2}{0.08^2+0.05^2}=0.72,\quad
K_\kappa=\frac{0.05^2}{0.05^2+0.30^2}=0.027.
\]

\textbf{Filter.} The update moves the real observed level and skew:
\[
m^+_\sigma=20.0+0.80(20.4-20.0)=20.32\%,\qquad
m^+_s=-0.35+0.72(-0.02)=-0.364.
\]
But it barely accepts the isolated curvature kink:
\[
m^+_\kappa=0.10+0.027(0.55-0.10)\approx0.112.
\]

\textbf{Persistence.} The unquoted wing is a different matter. For a deep-tail
functional $O_{\mathrm{tail}}$, observation precision is below requirement, so
the activation gate of \eqref{eq:persistence} is near one and the transported
prior supplies the tail anchor. The same day therefore has both effects:
Kalman filtering denoises the observed cluster, while prior persistence fills
the wing gap.
\end{example}

This example is deliberately small, but it captures the implementation rule: a
dense cluster with a bad internal residual is not a gap; it is a high-$R_t$
observation. A strike range with no reliable quotes is a gap; it belongs to
prior persistence.

%======================================================================
\section{Implementation plan}
\label{sec:implementation}

\subsection{State objects}

Add a pure numerical module, proposed as
\nolinkurl{backend/volfit/calib/observation_filter.py}, with immutable
dataclasses:
\begin{description}[leftmargin=1.7em,style=nextline]
\item[\texttt{FilterState}.]
Node key, as-of/source key, handle names, posterior mean $m^+$, covariance
$P^+$, timestamp, provenance, and reset reason.
\item[\texttt{FilterPrediction}.]
Predicted mean $m^-$, covariance $P^-$, transport distance, process-noise
breakdown.
\item[\texttt{FilterMeasurement}.]
Observed handles $z$, covariance $R$, quote-count/spread/RMS/inconsistency
breakdown.
\item[\texttt{FilterUpdate}.]
Innovation $z-Hm^-$, innovation covariance $S$, gain $K$, posterior mean and
covariance.
\end{description}

The numerical core should be pure numpy: no app state, no provider calls, no
calibration side effects.

\subsection{API and state integration}

An app layer, proposed as \nolinkurl{backend/volfit/api/observation_filter.py},
would:
\begin{enumerate}[leftmargin=1.6em]
\item Resolve the node key $(\mathrm{ticker},\mathrm{expiry},\mathrm{fitMode},
\mathrm{source},\mathrm{asOf})$.
\item Read prepared quotes and the latest data-only fit.
\item Build the measurement and covariance.
\item Read the previous filter state or seed from an active transported prior.
\item Apply the update and store diagnostics on \texttt{AppState}.
\item Expose a payload to the Smile and Graph views.
\end{enumerate}

The state must reset on source/as-of changes, expiry changes, fit-mode changes
that alter the observation loss, provider switches, large calendar gaps, and
manual quote edits that invalidate the measurement.

\subsection{Settings}

The first surfaced settings should be deliberately few:
\begingroup
\small
\begin{longtable}{p{0.38\textwidth}p{0.13\textwidth}p{0.41\textwidth}}
\caption{Proposed observation-filter settings.}\label{tab:settings}\\
\toprule Knob & Default & Role\\ \midrule
\endfirsthead \toprule Knob & Default & Role\\ \midrule \endhead
\texttt{observationFilterMode} & \texttt{off} &
\texttt{off} / \texttt{overlay} / \texttt{active}.\\
\texttt{observationFilterState} & ATM / skew / curv &
Handle coordinates carried by the filter.\\
\texttt{filterProcessVolBpSqrtDay} & $5$--$15$ bp &
Clock process noise for ATM level; skew/curvature scale separately.\\
\texttt{filterTransportNoiseScale} & $0.10$ &
Extra process noise per log-forward transport distance.\\
\texttt{filterResidualInflation} & on &
Use \eqref{eq:resid-inflation} to inflate $R_t$.\\
\texttt{filterMaxGain} & $1.0$ &
Optional safety cap; useful during pilot, removable later.\\
\texttt{filterResetHours} & session/EOD dependent &
Maximum gap before the state is reset rather than predicted.\\
\bottomrule
\end{longtable}
\endgroup

Hidden constants can mirror the graph precision scales: per-handle confidence,
precision floors/caps, spread half-life and quote-density saturation.

\subsection{Workflow placement}

The placement in the existing workflow is:
\[
\begin{aligned}
&\text{fetch snapshot}\to\text{prepare quotes}\to\text{data-only fit}\\
&\to\text{measurement }(z_t,R_t)
\to\text{predict/update filter}
\to\text{overlay or active MAP calibration}.
\end{aligned}
\]
For a background calibration job, this belongs between the prepared-quote plan
and the final fit commit. For live streaming, the filter may update every
snapshot while active calibration remains trigger-gated; overlay mode is useful
here because it shows whether the filter is tracking or lagging before it is
allowed to steer calibration.

%======================================================================
\section{Validation protocol}
\label{sec:validation}

The filter should not be accepted because charts look smooth. It needs a
temporal backtest that distinguishes denoising from signal damping.

\begin{enumerate}[leftmargin=1.6em]
\item \textbf{Consecutive snapshots.} Use stored chain snapshots with real
timestamps. Fit $t-1$, predict $t$, update on a thinned or noisy subset of
$t$, and score held-out quotes at $t$.
\item \textbf{Close-strike contradiction injection.} Perturb one or two close
strikes inside the spread model and verify that curvature noise is rejected
without suppressing level moves.
\item \textbf{Shock pass-through.} Raise the whole ATM cluster by a true
volatility jump with tight spreads. The filter must move with high gain.
\item \textbf{Gap preservation.} Remove the wings. The Kalman layer should not
invent tail information; prior persistence or graph reconstruction should carry
that load.
\item \textbf{Double-count guard.} Active mode must match the one-stage MAP
objective. A test should fail if the implementation applies a posterior target
and raw quotes as independent evidence.
\item \textbf{Calibration of uncertainty.} Standardized residuals,
$(\mathrm{heldout}-m^+)/\sqrt{P^++R_{\mathrm{heldout}}}$, should have mean near
zero and standard deviation near one, or conservatively below one.
\end{enumerate}

Success is not "lower every RMS." Success is lower held-out error in noisy
snapshots, lower shape jitter in repeated live refreshes, and no meaningful
lag on high-quality market moves.

%======================================================================
\section{Existing anchors and proposed additions}
\label{sec:anchors}

\begingroup
\small
\setlength{\tabcolsep}{3pt}
\begin{longtable}{@{}p{0.27\textwidth}p{0.32\textwidth}p{0.35\textwidth}@{}}
\caption{How the proposed filter should attach to today's code.}\\
\toprule
Responsibility & Existing anchor & Proposed addition\\
\midrule
\endfirsthead
\toprule Responsibility & Existing anchor & Proposed addition\\ \midrule
\endhead
Raw quote containers and snapshot timestamps &
\nolinkurl{backend/volfit/data/types.py} &
Use snapshot timestamp/source in the filter node key.\\
Quote preparation, de-Americanization and IV bands &
\nolinkurl{backend/volfit/api/quotes.py} &
Build measurements only after prepared quotes exist.\\
Bid--ask objective and band semantics &
\nolinkurl{backend/volfit/calib/band.py} &
Map inactive band rows into low measurement precision.\\
Calibration-consistent RMS &
\nolinkurl{backend/volfit/calib/rms.py} &
Use fit residuals for $R_t$ inflation.\\
Shared precision vocabulary &
\nolinkurl{backend/volfit/calib/precision.py},
\nolinkurl{backend/volfit/graph/precision.py} &
Reuse density, spread, freshness and caps for cheap $R_t$.\\
Prior-persistence builders and gates &
\nolinkurl{backend/volfit/calib/prior.py},
\nolinkurl{backend/volfit/calib/operators.py},
\nolinkurl{backend/volfit/calib/factors.py} &
Keep gap priors separate from the Kalman prediction prior.\\
Covariance-form Gaussian update pattern &
\nolinkurl{backend/volfit/graph/posterior.py} &
Mirror its marginal-variance and innovation diagnostics.\\
Workflow fetch/calibrate/stream placement &
\nolinkurl{backend/volfit/api/workflow.py},
\nolinkurl{backend/volfit/api/service.py} &
\nolinkurl{backend/volfit/api/observation_filter.py}.\\
Numerical filter core &
None today &
\nolinkurl{backend/volfit/calib/observation_filter.py}.\\
Backtest harness &
\nolinkurl{backend/backtest},
\nolinkurl{backend/tests/test_temporal_backtest.py} &
\nolinkurl{backend/backtest/observation_filter.py},
\nolinkurl{backend/tests/test_observation_filter.py}.\\
\bottomrule
\end{longtable}
\endgroup

%======================================================================
\section{What is original here}
\label{sec:original}

The Kalman equations are classical \cite{Kalman1960,AndersonMoore1979}. The
Vol-Fitter-specific contribution is the separation of three notions that are
usually blended in volatility tools:
\begin{enumerate}[leftmargin=1.6em]
\item \textbf{Quote likelihood}: today's prepared quotes, with bands, weights
and no-arbitrage model structure.
\item \textbf{Temporal filtering}: a transported latent smile state with
prediction and measurement covariance.
\item \textbf{Gap persistence}: a prior that activates only where current quotes
do not identify a functional.
\end{enumerate}
That separation is what makes the feature safe. The filter can denoise a noisy
observed smile without using yesterday as a fake quote, and prior persistence can
stabilize unobserved wings without damping a real current-market move.

\appendix
%======================================================================
\section{Hyperparameter atlas}
\label{app:atlas}

\begingroup
\small
\begin{longtable}{p{0.38\textwidth}p{0.13\textwidth}p{0.41\textwidth}}
\caption{Observation-filter hyperparameters, proposed v1.}\label{tab:atlas}\\
\toprule Knob & Default & Role\\ \midrule
\endfirsthead \toprule Knob & Default & Role\\ \midrule \endhead
\multicolumn{3}{l}{\emph{Surfaced}}\\ \midrule
\texttt{observationFilterMode} & \texttt{off} & Feature mode:
\texttt{off}, \texttt{overlay}, \texttt{active}.\\
\texttt{observationFilterState} & ATM / skew / curv & Handle set carried by the
filter.\\
\texttt{filterProcessVolBpSqrtDay} & $10$ bp & ATM process noise per square-root
calendar day.\\
\texttt{filterProcessSkewSqrtDay} & $0.02$ & Skew process noise scale.\\
\texttt{filterProcessCurvSqrtDay} & $0.05$ & Curvature process noise scale.\\
\texttt{filterTransportNoiseScale} & $0.10$ & Process-noise multiplier for
large spot/forward transports.\\
\texttt{filterResetHours} & source dependent & Reset rather than predict across
long data gaps.\\
\texttt{filterMaxGain} & $1.0$ & Pilot safety cap on diagonalized gains; should
not bind in normal operation.\\
\midrule
\multicolumn{3}{l}{\emph{Hidden / derived}}\\ \midrule
\texttt{HANDLE\_CONFIDENCE} & $[1,1,0.01]$ initially & Curvature is less
identified than level and skew, matching graph precision practice.\\
\texttt{RMS\_FLOOR} & existing precision floor & Prevents one excellent fit from
creating infinite observation precision.\\
\texttt{SPREAD\_HALF} & existing precision half-life & Wider bid--ask bands
inflate $R_t$.\\
\texttt{REF\_ATM\_QUOTES} & existing density saturation & Dense ATM support
raises level/skew precision.\\
\texttt{RESID\_INFLATION\_CAP} & TBD & Caps \eqref{eq:resid-inflation} during
the pilot to avoid one broken chain poisoning the state.\\
\texttt{SOURCE\_RESET\_POLICY} & strict & Reset on provider/as-of/source changes
unless an explicit bridge is implemented.\\
\bottomrule
\end{longtable}
\endgroup

%======================================================================
\section{Performance notes}
\label{app:perf}

\begin{enumerate}[leftmargin=1.6em]
\item \textbf{Overlay mode is cheap after calibration.} If a data-only fit has
already been computed, the Kalman update is a dense $d\times d$ solve with
$d\le6$ in v1. The cost is invisible next to calibration.
\item \textbf{The expensive part is covariance extraction.} Full
Jacobian-propagated $R_t$ requires either existing analytic Jacobians or a
Gauss--Newton approximation from the optimizer. The cheap precision builder
\eqref{eq:cheapR} is a sensible first overlay.
\item \textbf{Active MAP should not add a second fit pass by default.} The
prediction prior is just another residual block in the existing least-squares
stack. A diagnostic data-only prepass can improve $R_t$, but it should be an
explicit setting just like Note~13's prior prepass.
\item \textbf{State storage is small.} Per node, the state is $O(d^2)$: a mean,
covariance and a few diagnostics. Even thousands of nodes are negligible.
\item \textbf{Graph generalization is natural but later.} A graph Kalman filter
would replace the per-node $P_t^-$ with a block covariance over the universe.
That is Note~14's covariance-form posterior with time dynamics added; it should
come after the single-node filter is validated.
\end{enumerate}

%======================================================================
\section{Reference implementation}
\label{app:ref}

The numerical heart is small. The listing below is intentionally model-free: it
assumes the caller has already built the prediction and measurement moments.

\begin{lstlisting}[caption={Covariance-form Kalman update with Joseph covariance.}]
def kalman_update(mean_pred, cov_pred, obs, obs_cov, H=None):
    m = np.asarray(mean_pred, dtype=float)
    P = np.asarray(cov_pred, dtype=float)
    z = np.asarray(obs, dtype=float)
    R = np.asarray(obs_cov, dtype=float)
    H = np.eye(m.size) if H is None else np.asarray(H, dtype=float)

    innovation = z - H @ m
    S = H @ P @ H.T + R
    K = np.linalg.solve(S.T, (P @ H.T).T).T

    out_mean = m + K @ innovation
    I_KH = np.eye(m.size) - K @ H
    out_cov = I_KH @ P @ I_KH.T + K @ R @ K.T
    out_cov = 0.5 * (out_cov + out_cov.T)
    if np.any(np.linalg.eigvalsh(out_cov) < -1e-12):
        raise FloatingPointError("posterior covariance lost PSD")
    return out_mean, out_cov, innovation, S, K
\end{lstlisting}

The active calibration residual corresponding to \eqref{eq:active-map} is also
small once the model exposes a handle extractor:

\begin{lstlisting}[caption={Residual rows for the active MAP prediction prior.}]
def prediction_prior_residual(model_handles, mean_pred, cov_pred):
    L = np.linalg.cholesky(np.asarray(cov_pred, dtype=float))
    return np.linalg.solve(L, np.asarray(model_handles) - np.asarray(mean_pred))
\end{lstlisting}

In production the Cholesky solve should use a small jitter and report it; adding
jitter is a diagnostic event, not a silent repair.

\begin{thebibliography}{9}
\bibitem{Kalman1960} R.~E.~Kalman. A new approach to linear filtering and
prediction problems. \emph{Journal of Basic Engineering}, 82(1):35--45, 1960.
\bibitem{AndersonMoore1979} B.~D.~O.~Anderson and J.~B.~Moore.
\emph{Optimal Filtering}. Prentice-Hall, 1979.
\bibitem{Gelman2013} A.~Gelman et al. \emph{Bayesian Data Analysis}. CRC Press,
3rd ed., 2013.
\bibitem{Gatheral2006} J.~Gatheral. \emph{The Volatility Surface}. Wiley, 2006.
\end{thebibliography}

\end{document}
